Ensuring Fairness with Transparent Auditing of Quantitative Bias in AI Systems\\ \thanks{This project is partially supported by NSTC 112-2634-F-002-004 from National Science and Technology Council and AS-GCS-113-H04 from Academia Sinica.}

### 2024-06-18 paper/paper.tex
minor revision
Final version

\title{Ensuring Fairness with Transparent Auditing of Quantitative Bias in AI Systems\\
\thanks{Identify applicable funding agency here. If none, delete this.}
}

\author{\IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
}

@@

% \author{\IEEEauthorblockN{1\textsuperscript{st} Chih-Cheng Rex Yuan}
% \IEEEauthorblockA{\textit{Institute of Information Science} \\
% \textit{Academia Sinica}\\
% Taipei, Taiwan \\
% hello@rexyuan.com}
% \and
% \IEEEauthorblockN{2\textsuperscript{nd} Bow-Yaw Wang}
% \IEEEauthorblockA{\textit{Institute of Information Science} \\
% \textit{Academia Sinica}\\
% Taipei, Taiwan \\
% bywang@iis.sinica.edu.tw}
% }

@@

AI system, however, may be biased. Factors such as inherent biases in original data sets or flaws in algorithm designs could contribute to bias within AI systems. When left unchecked, the ramifications of such biases extend far beyond mere inaccuracies; they can lead to devastating consequences, such as amplifying systemic injustices, perpetuating group discrimination, and exacerbating societal inequalities.

@@

Our tool builds upon transparent definitions widely accepted within the scientific community. By providing an abstraction layer atop these definitions, we enhance their flexibility and transparency, enabling users to adapt them to diverse contexts with ease and reveal them to the public. Moreover, we prioritize transparency throughout our tool's development process, offering access to the source code for scrutiny, thereby ensuring transparency from conception to implementation.

Our open-sourced tool is written in Python and is offered as a Python package. It supports common dataset formats such as CSV. The public domain source code is available at \url{https://github.com/RexYuan/Shu}.

@@

\subsection{Preliminaries}
For the following definition, we will use $Y$ to denote the ground truth of an outcome; $\hat{Y}$ to denote the predicated result of an outcome; $Y = 1$ and $\hat{Y} = 1$ to denote them being accepted or positive. For example, let $Y$ be recidivism. Then $Y = 1$ means the case of an individual actually recidivating and $\hat{Y} = 0$ means the prediction of an individual recidivating is negative. In addition, we use $V$ and $\hat{V}$ when the truth and the prediction aren't binary. For example, we denote the COMPAS score by $\hat{V}$, which ranges from 1 to 10.

@@

A \emph{score predicate} of rows $\sco : \mathtt{row} \implies \alpha$ takes a row and gives its predicted score. Similar to positive predicate, a score predicate of model results $\scom : \alpha \implies \beta$ takes a model's result and gives its score.

@@

A \emph{legitimate predicate} $\leg : \prod_i \alpha_i \implies \mathtt{row} \implies \mathtt{bool}$ takes $n$ parameters and returns a row predicate. For example, $\leg(x)(r_i) := \mathtt{int}(r_i(\text{``priors\_count''})) > x$ first take a parameter $x$ and then decides if priors count is larger than it.

@@

\subsection{Definitions}
We abstracted the idea of privileged groups and positive prediction as predicates to maximize flexibility. With the proper predicates, any model output can be used; even prose-like responses of generative models can be included given adequate predicates.

@@

For equal opportunity, recall its formal definition:
\begin{align*}
    \abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} \leq \epsilon
\end{align*}
To model the $Y$, $\hat{Y}$, and $S$, we define the corresponding $\tru$, $\pos$, and $\priv$. We have $Y = 1$ if and only if $\tru$ is true; $\hat{Y} = 1$ if and only if $\pos$ is true; $S = 1$ if and only if $\priv$ is true; and vice versa. This way, we can calculate equal opportunity as a function:
\begin{align*}
    \text{equal\_opportunity} (\epsilon, \model, \priv, \pos, \tru)
\end{align*}

@@

We will set the unprivileged predicate be to ``African-American'' so that the unprivileged group will be African-American to test if the African-American race is discriminated.
\[
    \priv(r_i) := r_i(\text{``race''}) \neq \text{``African-American''}
\]

As ProPublica referenced Northpointe's COMPAS Practitioners Guide and cited that ``medium''(5-7) and ``high''(8-10) categories of scores are considered to indicate a risk of recidivism, we set the positive predicate using the readily available category.
\[
    \pos(r_i) := r_i(\text{``score\_text''}) \in \set{\text{``Medium''},\text{``High''}}
\]
On the other hand, when we're taking prediction input from our makeshift model, we simply return its result because it already gives a binary $\mathtt{bool}$ result: $\posm(r_i) := \model(r_i)$

@@

\begin{figure*}[h]
    \centering
    \begin{minipage}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{Non-African-American}
        \subcaption{Privileged Group: African-American}\label{fig:1}
    \end{minipage}
    \hfill
    \begin{minipage}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{Non-Caucasian}
        \subcaption{Privileged Group: Caucasian}\label{fig:2}
    \end{minipage}
    \hfill
    \begin{minipage}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{Non-Asian}
        \subcaption{Privileged Group: Asian}\label{fig:2}
    \end{minipage}
    \hfill
    \begin{minipage}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{Non-Native American}
        \subcaption{Privileged Group: Native American}\label{fig:2}
    \end{minipage}
    \hfill
    \begin{minipage}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{Non-Hispanic}
        \subcaption{Privileged Group: Hispanic}\label{fig:2}
    \end{minipage}
    \hfill
    \begin{minipage}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{race balance}
        \subcaption{Balance}\label{fig:2}
    \end{minipage}
    \caption{Fairness measures of ProPublica COMPAS dataset grouped by race}
    \label{fig:compas}
\end{figure*}

\begin{figure*}[h]
    \centering
    \begin{minipage}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{Non-Less than 25}
        \subcaption{Privileged Group: Greater than 25}\label{fig:1}
    \end{minipage}
    \hfill
    \begin{minipage}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{Male}
        \subcaption{Privileged Group: Male}\label{fig:2}
    \end{minipage}
    \hfill
    \begin{minipage}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{Non-25 - 45}
        \subcaption{Privileged Group: Less than 25 or Greater than 45}\label{fig:2}
    \end{minipage}
    \hfill
    \begin{minipage}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{Misdemeanor}
        \subcaption{Privileged Group: Misdemeanor}\label{fig:2}
    \end{minipage}
    \hfill
    \begin{minipage}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{Non-Greater than 45}
        \subcaption{Privileged Group: Less than 45}\label{fig:2}
    \end{minipage}
    \hfill
    \begin{minipage}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{misc balance}
        \subcaption{Balance}\label{fig:2}
    \end{minipage}
    \caption{Fairness measures of ProPublica COMPAS dataset grouped by age, sex, and charged degree}
    \label{fig:compas2}
\end{figure*}

>>>>

\title{Ensuring Fairness with Transparent Auditing of Quantitative Bias in AI Systems\\
\thanks{This project is partially supported by NSTC 112-2634-F-002-004 from National Science and Technology Council and AS-GCS-113-H04 from Academia Sinica.}
}

@@

\author{\IEEEauthorblockN{Chih-Cheng Rex Yuan}
\IEEEauthorblockA{\textit{Institute of Information Science} \\
\textit{Academia Sinica}\\
Taipei, Taiwan \\
hello@rexyuan.com}
\and
\IEEEauthorblockN{Bow-Yaw Wang}
\IEEEauthorblockA{\textit{Institute of Information Science} \\
\textit{Academia Sinica}\\
Taipei, Taiwan \\
bywang@iis.sinica.edu.tw}
}

@@

AI systems, however, may be biased. Factors such as inherent biases in original data sets or flaws in algorithm designs could contribute to bias within AI systems. When left unchecked, the ramifications of such biases extend far beyond mere inaccuracies; they can lead to devastating consequences, such as amplifying systemic injustices, perpetuating group discrimination, and exacerbating societal inequalities.

@@

Our tool builds upon transparent definitions widely accepted within the scientific community. By providing an abstraction layer atop these definitions, we enhance their flexibility and transparency, which enables users to adapt them to diverse contexts with ease and reveal them to the public. Moreover, we prioritize transparency throughout our tool's development process, offering access to the source code for scrutiny, thereby ensuring transparency from conception to implementation.

Our open-sourced tool is written in Python and is offered as a Python package. It supports common dataset formats such as CSV. The package is available at \url{https://pypi.org/project/fairness-checker/}.

@@

\subsection{Preliminaries}
For the following definition, we will use $Y$ to denote the ground truth of an outcome; $\hat{Y}$ to denote the predicated result of an outcome; $Y = 1$ and $\hat{Y} = 1$ to denote them being accepted or positive. For example, let $Y$ be recidivism. Then $Y = 1$ means the case of an individual actually recidivating and $\hat{Y} = 0$ means the prediction of an individual recidivating is negative. In addition, we use $V$ and $\hat{V}$ when the truth and the prediction are not binary. For example, we denote the COMPAS score by $\hat{V}$, which ranges from 1 to 10.

@@

A \emph{score predicate} of rows $\sco : \mathtt{row} \implies \beta$ takes a row and gives its predicted score of some type $\beta$. Similar to positive predicate, a score predicate of model results $\scom : \alpha \implies \beta$ takes a model's result and gives its score.

@@

A \emph{legitimate predicate} $\leg : \prod_i \alpha_i \implies \mathtt{row} \implies \mathtt{bool}$ takes $n$ parameters and returns a row predicate. For example, $\leg(x)(r_i) := \mathtt{int}(r_i(\text{``priors\_count''})) > x$ first takes a parameter $x$ and then decides if priors count is larger than it.

@@

\subsection{Definitions}
We abstracted the idea of privileged groups and positive prediction as predicates to maximize flexibility. With proper predicates, any model output can be used; even prose-like responses of generative models can be included given adequate predicates.

@@

For equal opportunity, recall its formal definition:
\begin{align*}
    \abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} \leq \epsilon
\end{align*}
To model $Y$, $\hat{Y}$, and $S$, we define the corresponding $\tru$, $\pos$, and $\priv$. We have $Y = 1$ if and only if $\tru$ is true; $\hat{Y} = 1$ if and only if $\pos$ is true; $S = 1$ if and only if $\priv$ is true; and vice versa. This way, we can calculate equal opportunity as a function:
\begin{align*}
    \text{equal\_opportunity} (\epsilon, \model, \priv, \pos, \tru)
\end{align*}

@@

We will set the unprivileged predicate to be ``African-American'' so that the unprivileged group will be African-American. Then we test if the African-American race is discriminated.
\[
    \priv(r_i) := r_i(\text{``race''}) \neq \text{``African-American''}
\]

As ProPublica referenced Northpointe's COMPAS Practitioners Guide and cited that ``medium''(5-7) and ``high''(8-10) categories of scores are considered to indicate a risk of recidivism, we set the positive predicate using the readily available category.
\[
    \pos(r_i) := r_i(\text{``score\_text''}) \in \set{\text{``Medium''},\text{``High''}}
\]

@@

\begin{figure*}[h]
    \centering
    \begin{minipage}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{African-American}
        \subcaption{Unprivileged Group: African-American}\label{fig:1}
    \end{minipage}
    \hfill
    \begin{minipage}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{Caucasian}
        \subcaption{Unprivileged Group: Caucasian}\label{fig:2}
    \end{minipage}
    \hfill
    \begin{minipage}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{Asian}
        \subcaption{Unprivileged Group: Asian}\label{fig:2}
    \end{minipage}
    \hfill
    \begin{minipage}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{Native American}
        \subcaption{Unprivileged Group: Native American}\label{fig:2}
    \end{minipage}
    \hfill
    \begin{minipage}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{Hispanic}
        \subcaption{Unprivileged Group: Hispanic}\label{fig:2}
    \end{minipage}
    \hfill
    \begin{minipage}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{race balance}
        \subcaption{Balance}\label{fig:2}
    \end{minipage}
    \caption{Fairness measures of ProPublica COMPAS dataset grouped by race}
    \label{fig:compas}
\end{figure*}

\begin{figure*}[h]
    \centering
    \begin{minipage}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{Less than 25}
        \subcaption{Unprivileged Group: Less than 25}\label{fig:1}
    \end{minipage}
    \hfill
    \begin{minipage}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{Male}
        \subcaption{Privileged Group: Male}\label{fig:2}
    \end{minipage}
    \hfill
    \begin{minipage}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{25 - 45}
        \subcaption{Unprivileged Group: 25 - 45}\label{fig:2}
    \end{minipage}
    \hfill
    \begin{minipage}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{Misdemeanor}
        \subcaption{Privileged Group: Misdemeanor}\label{fig:2}
    \end{minipage}
    \hfill
    \begin{minipage}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{Greater than 45}
        \subcaption{Unprivileged Group: Greater than 45}\label{fig:2}
    \end{minipage}
    \hfill
    \begin{minipage}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{misc balance}
        \subcaption{Balance}\label{fig:2}
    \end{minipage}
    \caption{Fairness measures of ProPublica COMPAS dataset grouped by age, sex, and charged degree}
    \label{fig:compas2}
\end{figure*}
